The Host Galaxies of Rapidly Evolving Transients in the Dark Energy Survey

### 2020-03-10 RET_Hosts.tex
Updates from Overleaf

 Long duration gamma-ray bursts (LGRBs), although first discovered in the 1960s \citep{Klebesadel1973} were only unequivocally linked to collapsing massive stars through their associations with broad-lined type Ic SNe \citep{Galama1998,Hjorth2003}. Thought to be caused by accretion onto a newly-formed black hole at the centre of a collapsing, rapidly-rotating massive star \citep[e.g.][]{Woosley1993,Woosley2006a,Woosley2006b}, LGRBs comprise roughly $1\%$ of all SNe Ic, themselves making up only 15\% of all CCSNe \citep{Kelly2012,Graham2016}. The second exotic class of SNe is the particularly bright superluminous supernovae (SLSNe; e.g. \citealt{Quimby2011, Gal-Yam2012}). Originally grouped due to their slowly-evolving lightcurves and extreme luminosity (peaking at $M_B < -21$~mag; 10-100 times brighter than regular CCSNe), recent observations have revealed a continuum of spectroscopically similar objects with peaks as faint as $M_B \sim -19$~mag \citep{DeCia2018,Lunnan2018,Angus2019}, similar to the bright end of the CCSN luminosity function \citet{Li2011,Grayling2020}. The lightcurve evolution of SLSNe is not well described by models of Ni-56 decay, with the most popular alternative hypothesis being the magnetic coupling of the ejecta with the spin down of a newly formed, rapidly rotating magnetar.

@@

\begin{figure*}
\includegraphics[width=0.7\textwidth]{figs/RET_Mosaic.png}
\caption{Selection of DES RET host galaxies in an RGB composite of the DES $gri$ band deep coadds from \citet{Wiseman2020}. The locations of the transients are indicated with cyan crosses. The stamps have a size of 10\arcsec in each direction.
\label{fig:mag_dist}}
\end{figure*}

@@

\miika{Miika to verify!}
The original method of finding RETs in the DES-SN data (and presented in \citetalias{Pursiainen2018}) was designed to be simple and used light curve modelling with Gaussian and linear fits. The simplistic method made it possible to look for exotic transients without knowing their observed characteristics beforehand and resulted in a large sample of photometrically selected fast transients. However, as the search was simplistic and relied heavily on visual inspection of the available data (etc. images, light curves, host galaxy information), it is impossible to be sure how complete the sample is. For instance, due to the large redshift range within the sample it is entirely possible that distant events could have been missed due to time dilation. Here, a more sophisticated search method is presented. As only a fraction of transients in DES-SN have redshift information from host galaxy, the search must be done in the observer frame.  The key features of RETs that separate them from most traditional SNe types are the fast light curve evolution and blue colour at peak. Even though both of these quantities depend on the redshift of the transient, they still seem to distinguish the the fast events from traditional SN types.

@@

To compare with CCSNe, we draw on the sample of SNe II from PTF \citep{Stoll2013} and the compilation of untargeted SESNe from \citet{Sanders2012}. We use the sample of GRB host galaxies of \citet{Kruehler2015}, using only galaxies with $z<1$ in order to maintain completeness. To investigate similarities with SLSNe, we use the PTF sample of \citet{Perley2016c}.

@@

As for SFR, we show the cumulative distribution of sSFR in Fig.~\ref{fig:ssfr_cum}. The RET hosts are clearly shifted to higher sSFRs than CCSNe. To statistically compare the host sSFR distribution of RETs with the other samples, we employ the method of \citet{Wiseman2020}. For each pair of samples, we model the PDFs as skewed normal distributions. For both samples, we use identical normal priors for the `mean' and `scale'\footnote{See \citet{Wiseman2020} for a detailed description of the parameters describing the skewed normal distributions}, centred on the mean and twice the standard deviation of the two samples combined. The results from the simultaneous fitting of the samples reveals:

@@

We compare the metallicity distributions in the same way as the sSFRs. The RET host metallicity distribution is significantly different to the SNe II and SDSS distributions, with no overlap in the posterior distributions for the skewed-normal distribution peak and skewness. On the other hand, simultaneous fits with SESNe show very similar distributions.

@@

The sample of DES RETs shows a preference for low-metallicity, strongly star-forming host environments. The PDF of their metallcities displays a strong similarity to the hosts of SNe Ibc, as well as LGRBs. There is a clear difference to the PDFs of SNe II, which follow SDSS field galaxies. The preference for low-metallicity systems is not as strong as for LGRBs or SLSNe, but the highest metallicities found in all three samples are very similar at around solar metallicity. This result is further suggestive of a stripped-envelope, massive-star origin for RETs. 
The population of RET hosts lies, on average, between CCSNe and LGRBs/SLSNe in terms of both star formation and metallicity. A loose correlation exists between the luminosity and rarity of events, and the host galaxy conditions required for their formation. The rough rate of RETs ($\geq 10^{-6} \mathrm{Mpc}^{-3} \mathrm{yr}^{-1}$), (\citetalias{Pursiainen2018}) is $\sim1\%$ of the CCSN rate \citep{Li2011,Horiuchi2011}, which itself is divided into the more common SNe II and sub-dominant SESNe \citep{Kelly2012,Frohmaier2020}. At $\sim1\%$ of the CCSN rate, RETs are more common than SLSNe ($\sim0.01 - 0.05\%$ of CCSNe; \citealt{McCrum2015,Prajz2016,Frohamier2020}) and LGRBs (intrinsically $\sim0.08\%$ when accounting for beaming; \citealt{Graham2016}). These figures place the DES RETs between extreme objects (SLSNe, LGRBs) and more common SNe (SNe II, SESNe) in terms of rate, matching their location in host galaxy parameter space. While stressing that these associations are lose - rates are uncertain and host galaxy parameters span wide ranges for all transients - they are both linked to the respective transients' progenitor channels. Based upon both indicators, it is reasonable to infer that RETs are linked to very massive stars that also require some extreme properties such as rapid rotation, albeit not as much as progenitors as SLSNe or LGRBs. It could therefore be possible that RETs are an intermediate and/or precursory step, whereby the initial collapse of the star occurs leading to a shocked photosphere, but conditions are not highly tuned enough for a LGRB or SLSN and the respective central engine does not form.

>>>>

 Long duration gamma-ray bursts (LGRBs), although first discovered in the 1960s \citep{Klebesadel1973} were only unequivocally linked to collapsing massive stars through their associations with broad-lined type Ic SNe \citep{Galama1998,Hjorth2003}. Thought to be caused by accretion onto a newly-formed black hole at the centre of a collapsing, rapidly-rotating massive star \citep[e.g.][]{Woosley1993,Woosley2006a,Woosley2006b}, LGRBs comprise roughly $1\%$ of all SNe Ic, themselves making up only 15\% of all CCSNe \citep{Kelly2012,Graham2016}. The second exotic class of SNe is the particularly bright superluminous supernovae (SLSNe; e.g. \citealt{Quimby2011, Gal-Yam2012}). Originally grouped due to their slowly-evolving lightcurves and extreme luminosity (peaking at $M_B < -21$~mag; 10-100 times brighter than regular CCSNe), recent observations have revealed a continuum of spectroscopically similar objects with peaks as faint as $M_B \sim -19$~mag \citep{DeCia2018,Lunnan2018,Angus2019}, similar to the bright end of the CCSN luminosity function \citet{Li2011}. The lightcurve evolution of SLSNe is not well described by models of Ni-56 decay, with the most popular alternative hypothesis being the magnetic coupling of the ejecta with the spin down of a newly formed, rapidly rotating magnetar.

@@

\begin{figure*}
\includegraphics[width=0.7\textwidth]{figs/RET_Mosaic.png}
\caption{Selection of DES RET host galaxies in an RGB composite of the DES $gri$ band deep coadds from \citet{Wiseman2020}. The locations of the transients are indicated with cyan crosses. The stamps have a size of 10\arcsec in each direction.
\label{fig:mosaic}}
\end{figure*}

@@

The original method of finding RETs in the DES-SN data (and presented in \citetalias{Pursiainen2018}) was designed to be simple and used light curve modelling with Gaussian and linear fits. The simplistic method made it possible to look for exotic transients without knowing their observed characteristics beforehand and resulted in a large sample of photometrically selected fast transients. However, as the search was simplistic and relied heavily on visual inspection of the available data (etc. images, light curves, host galaxy information), it is impossible to be sure how complete the sample is. For instance, due to the large redshift range within the sample it is entirely possible that distant events could have been missed due to time dilation. Here, a more sophisticated search method is presented. As only a fraction of transients in DES-SN have redshift information from host galaxy, the search must be done in the observer frame.  The key features of RETs that separate them from most traditional SNe types are the fast light curve evolution and blue colour at peak. Even though both of these quantities depend on the redshift of the transient, they still seem to distinguish the the fast events from traditional SN types.

@@

In compiling a set of comparison samples, we aim for the least biased selections possible. This requires surveys to be untargetted (they were not monitoring certain galaxies in order to search for SNe), and ideally complete to a certain volume limit. While in practice the second of these criteria is difficult to achieve, we are able to choose comparison samples from untargetted surveys to mitigate initial selection biases.

To compare with CCSNe, we draw on the untargetted sample of 47 SNe II from PTF \citep{Stoll2013}, which is likely complete in terms of hosts (all SNe have an associated host). While this sample lies at much lower redshift than the DES RETs (a maximum of 0.18 and mean of 0.05), redshift evolution is easier to account for than unknown incompleteness. We add to this the compilation of 56 untargeted SESNe from \citet{Sanders2012}, with a maximum redshift of 0.26 and a mean of 0.05. Since \citet{Sanders2012} do not report host galaxy magnitudes, stellar masses or SFRs, we cross-correlated the SN positions with the the Sloan Digital Sky Survey (SDSS; \citealt{York2000}) Data Release 16 (DR16; \citealt{Ahumada2019}) and perform our own SED fit using the method outlined in Section \ref{subsec:sedfit}.

We use the sample of GRB host galaxies of \citet{Kruehler2015}, using only galaxies with $z<1$ in order to maintain completeness, resulting in a sample of 29 hosts with a mean redshift of 0.66. To investigate similarities with SLSNe, we use the PTF sample of \citet{Perley2016c} with a mean redshift of 0.24 and a maximum of 0.50.

The host galaxy properties of the above samples are not all derived using the same methods. In terms of SED fitting, the largest systematic offsets in derived properties are due to differences in the assumed initial mass function (IMF). \citet{Stoll2013} and \citet{Drout2014} assume a \citet{Salpeter1955} IMF whereas all other samples considered (including those calculated in this work in Section \ref{subsec:sedfit}) are determined assuming a \citet{Chabrier2003} IMF. Stellar masses and star-formation rates derived using a Salpeter IMF are believed to be roughly $\sim 0.58$ \citep{Speagle2014}, and we convert by this factor in order to compare them.

@@

As for SFR, we show the cumulative distribution of sSFR in Fig.~\ref{fig:ssfr_cum}. The RET hosts are clearly shifted to higher sSFRs than CCSNe. To statistically compare the host sSFR distribution of RETs with the other samples, we employ the method of \citet{Wiseman2020}. For each pair of samples, we model the PDFs as skewed normal distributions. For both samples, we use identical normal priors for the `loc' (location, i.e. central value, identical to mean for a skewness of 0) and `scale' (spread, identical to standard deviation for a skewness of 0)\footnote{See \citet{Wiseman2020} for a detailed description of the fitting procedure and the parameters describing the skewed normal distributions}, centred on the mean and twice the standard deviation of the two samples combined. The results from the simultaneous fitting shows RET hosts to be shifted to higher sSFRs CCSNe. In 100\% of the posterior samples the RETs were located at higher average sSFRs than SNe II, while the same was true 76\% of the time for RETs when comparing with SESNe.

RET host galaxies are significantly lower in sSFR than LGRBs. While the distributions are similar in shape, with a mean difference in scale of 0.28, the mean difference in loc is -1.64, with no overlap between the posterior distributions. The sSFR distribution of SLSNe hosts is heavily positively skewed (with alpha values of $5.54^{+1.8}_{-0.98}$ compared to $-0.18^{+0.75}_{-0.76}$ for RET hosts). The strong high-sSFR tail shows SLSNe to derive from a different population to RETs.

@@

We compare the metallicity distributions in the same way as the sSFRs. The RET host metallicity distribution shows a broad peak, leading to two families of skewed-Gaussians that fit it well, one with a low-valued centre  ($12 + \log \mathrm{(O/H)} \sim 8.1$) and a positive skew, and the other with a higher-valued centre ($12 + \log \mathrm{(O/H)} \sim 8.6$). Comparing the DES RETs to the \citet{Stoll2013} SNe II shows the latter to be centred around 8.8, with the centre being greater than the RETs in 94\% of samples. We determine that the RET host metallicities are derived from a different population to the SNe II. On the other hand, simultaneous fits with SESNe show very similar distributions, including a smaller higher-metallicity peak, such that they are indistinguishable statistically.

The median CDFs of LGRBs and SLSNe show divergence from the RETs, particularly at low metallicity. As a result, the central locations of their fits are shifted compared to the RETs. In 67\% of samples, the RET sample had a higher central metallicity than the LGRBS, with RETs also showing a broader distribution 67\% of the time. While these effects are not as significant as with the RETs - SNe II comparison, there is mild evidence that RETs located in galaxies with higher metal content than LGRBs. The effect is more pronounced for SLSNe, where the RETs have a higher metallicity for the distribution peak 92\% of the time. The SLSN distribution is also more strongly skewed, with 89\% of the posterior distribution being more strongly skewed than the RETs. There is thus mild-to-strong evidence that RETs occur in more metal-rich environments than SLSNe.

@@

The sample of DES RETs shows a preference for low-metallicity, strongly star-forming host environments. The PDF of their metallcities displays a strong similarity to the hosts of SNe Ibc, as well as LGRBs. There is a clear difference to the PDFs of SNe II, which follow SDSS field galaxies. The preference for low-metallicity systems is not as strong as for LGRBs or SLSNe, but the highest metallicities found in all three samples are very similar at around solar metallicity. This result is further suggestive of a stripped-envelope, massive-star origin for RETs. 
The population of RET hosts lies, on average, between CCSNe and LGRBs/SLSNe in terms of both star formation and metallicity. A loose correlation exists between the luminosity and rarity of events, and the host galaxy conditions required for their formation. The rough rate of RETs ($\geq 10^{-6} \mathrm{Mpc}^{-3} \mathrm{yr}^{-1}$), (\citetalias{Drout2014,Pursiainen2018,Tampo2020}) is $\sim1\%$ of the CCSN rate \citep{Li2011,Horiuchi2011,Strolger2015}, which itself is divided into the more common SNe II and sub-dominant SESNe \citep{Kelly2012,Frohmaier2020}. At $\sim1\%$ of the CCSN rate, RETs are more common than SLSNe ($\sim0.01 - 0.05\%$ of CCSNe; \citealt{McCrum2015,Prajs2017,Frohamier2020}) and LGRBs (intrinsically $\sim0.08\%$ when accounting for beaming; \citealt{Graham2016}). These figures place the DES RETs between extreme objects (SLSNe, LGRBs) and more common SNe (SNe II, SESNe) in terms of rate, matching their location in host galaxy parameter space. While stressing that these associations are lose - rates are uncertain and host galaxy parameters span wide ranges for all transients - they are both linked to the respective transients' progenitor channels. Based upon both indicators, it is reasonable to infer that RETs are linked to very massive stars that also require some extreme properties such as rapid rotation, albeit not as much as progenitors as SLSNe or LGRBs. It could therefore be possible that RETs are an intermediate and/or precursory step, whereby the initial collapse of the star occurs leading to a shocked photosphere, but conditions are not highly tuned enough for a LGRB or SLSN and the respective central engine does not form.
